NLTK: The Natural Language Toolkit

### 2002-03-21 technical/acl2002/submission.tex
- Added section on documentation - General clean-up sweep - Commented out abstract
svn/trunk@519

\documentclass[11pt]{article}
\usepackage{acl02sub,url,alltt,epsfig}
\title{NLTK: The Natural Language Toolkit}
\author{
Edward Loper and Steven Bird\\
Department of Computer and Information Science and Linguistic Data Consortium\\
University of Pennsylvania, Philadelphia, PA 19104-6389, USA
}
\summary{%
NLTK, the Natural Language Toolkit, is a suite of program modules,
tutorials and problem sets, providing ready-to-use CL courseware.
NLTK covers symbolic and statistical
NLP, and is interfaced to annotated corpora.
Students augment and replace existing components, learning
structured programming by example, and manipulating sophisticated
models from the outset.}
\paperid{Pxxxx}
\keywords{teaching, courseware, Python, corpora}
\contact{Edward Loper}
\conference{this paper has not been submitted to any other conferences}
\date{\today}

@@

% Outline:
%   - abstract
%   - introduction
%   - choice of programming language
%   - design criteria
%   - python implementation (??)
%   - modules
%   - uses
%     - assignments/projects
%     - class demonstrations
%   - evaluation
%   - other approaches
%   - conclusion

% To do:
%   - Fill in OVERVIEW in the intro
%   - Write up evaluation
%   - Write up other approaches
%   - Write up conclusion
%   - Get rid of Python implementation section, or at least shorten it?
%   - General clean-up
%   - ispell

% We need more content on:
%   - documentation
%   - corpora?

@@

\begin{abstract}
Students in computational linguistics courses must often learn a new
programming language.  In some cases, such as when courses are offered
in linguistics departments, the students may be learning to program
for the very time.  In order to do interesting projects, it is usually
necessary for students to do many low-level ``housekeeping'' tasks.
At the same time, teachers of computational linguistics courses
sometimes feel that they spend too much time teaching students to
program, and not enough time teaching the subject itself.  They may
even avoid programming assignments altogether.  However, we believe
that it is crucial for a first computational linguistics course to
include a strong practical component, in which students develop real
programs to solve real problems with real data.

Python is a new object-oriented scripting language which runs on all
platforms.  Python has been praised as ``executable pseudocode,'' since
programs are so easy to write.  Recently, we have been developing
NLTK, an open-source Natural Language Toolkit written in Python.
In this presentation, we will motivate, describe and demonstrate NLTK.

NLTK, the Natural Language Toolkit, is a suite of program modules,
tutorials and problem sets.  NLTK covers symbolic and statistical
natural language processing, and is interfaced to annotated corpora.
Students augment and replace existing NLTK components, learning
structured programming by example, and manipulating sophisticated
models from the outset.  Along with extensive documentation and
problem sets, NLTK provides self-contained, ready-to-use CL
courseware.
\end{abstract}

\section{Introduction}

Teachers of introductory courses on computational linguistics are
often faced with the challenge of setting up a practical programming
component for student assignments and projects.  This is difficult not
just because of the variety of data structures, but also because of
the diverse range of topics which may need to be included in the
syllabus.  A widespread practice is to teach multiple programming
languages, where each language provides native data structures and functions
that are a good fit for the task at hand (e.g. Prolog for parsing,
Perl for corpus processing, a finite-state toolkit for morphological
analysis).  By relying on the built-in features of various languages,
the teacher avoids having to develop a lot of software infrastructure.
An unfortunate consequence is that a significant part of the course
must be devoted to teaching these languages.  Further, many
interesting projects require the languages to be bridged, e.g. a
project that involved syntactic parsing of corpus data from a
morphologically rich language would involve all three languages
mentioned above (Perl for file I/O, format conversions and overall
program control, with calls out to a finite state toolkit for
morphological analysis and to a Prolog engine for parsing).  It is
clear that these considerable overheads and shortcomings warrant a
fresh approach.

Apart from the practical component, computational linguistics courses
may also depend on software for in-class demonstrations.  This context
calls for highly interactive graphical user interfaces making it
possible to view program state (e.g. the chart of a chart parser),
observe program execution step-by-step (e.g. execution of a
finite-state machine), and even make minor modifications to programs
in response to ``what if'' questions from the class.  Because of these
difficulties it is common to avoid live demonstrations, and keep
classes for theoretical presentations only.  Apart from being dull,
this approach leaves students to solve important practical problems on
their own, or to deal with them less efficiently in office hours.

In this paper we describe a new approach to the above challenges,
a streamlined and flexible way of organizing the practical component
of an introductory computational linguistics course.  We describe
NLTK, the Natural Language Toolkit, which we have developed in
conjunction with a course we have taught at the University of Pennsylvania.

OVERVIEW OF THE PAPER

@@

\section{Choice of programming language}

The most basic step in setting up a practical component is choosing a
suitable programming language.  Here we list the desiderata that influence
our choice.

First, the language must have a shallow learning curve, so that novice
programmers get immediate rewards for their efforts.
Second, the language must support rapid prototyping and a short develop/test cycle;
an obligatory compilation step is a serious detraction.
Third, the code should be self-documenting,
with a transparent syntax and semantics.
Fourth, it should be easy to write structured programs, ideally object-oriented
but without the burden associated with languages like C++.
Finally, the language must have an easy-to-use graphics library to support
the development of simple graphical user interfaces.

In surveying the available languages, we believe that Python offers an especially good
fit to the above requirements.  Python is an object-oriented scripting language
developed by Guido van Rossum [REF] and available on all platforms
\url{www.python.org}.
Python offers a shallow learning curve by virtue of the
fact that it was designed to be easily learnt by children [REF].
As an interpreted language, Python is suitable for rapid prototyping.
Python code is exceptionally readable, and it has been praised as
``executable pseudocode.''
Python is an object-oriented language, but not punitively so, and it
is easy to encapsulate data and methods inside Python classes.
Finally, Python has an interface to the ``tk'' graphics toolkit, and
writing graphical interfaces is straightforward.

\section{Design criteria}

Several criteria were considered in the design and implementation of
the toolkit.  The design criteria are divided into primary and
secondary criteria, and within those sections are listed in the order
of their importance.  It was also important to decide what goals the
toolkit would \emph{not} attempt to accomplish; we therefore include
an explicit set of non-requirements, which the toolkit is not expected
to satisfy.

@@

\noindent\textit{\textbf{Ease of Use.}} The primary purpose of the toolkit is
to allow students to concentrate on building NLP systems.  The more
time students must spend learning to use the toolkit, the less useful
it is.

\noindent\textit{\textbf{Consistency.}} The toolkit should use consistent data
structures and interfaces.

\noindent\textit{\textbf{Extensibility.}} The toolkit should easily
accommodate new components, whether those components replicate or
extend the toolkit's existing functionality.  Thus, the toolkit's
design should be modular, with simple and well-defined interfaces
between modules.  The toolkit should also be structured in such a way
that it is obvious where new extensions would fit into the toolkit's
infrastructure.

\noindent\textit{\textbf{Documentation.}} The toolkit, its data structures,
and its implementation all need to be documented.  Three documentation
types are necessary: technical reports, to explain the toolkit's
design and implementation; tutorials, to teach students how to use the
toolkit to perform specific tasks; and reference documentation, to
describe every module, interface, class and method.  All nomenclature
must be carefully chosen and consistently used.

@@

\noindent\textit{\textbf{Simplicity.}} The toolkit should structure the
complexities of building NLP systems, not hide them.  Therefore, each
class defined by the toolkit should be simple enough that a student
could implement it by the time they finish the course.

\noindent\textit{\textbf{Modularity.}} The interaction between different
components of the toolkit should be kept to a minimum.  In particular,
it should be possible to complete individual projects using small
parts of the toolkit, without worrying about how they interact with
the rest of the toolkit.  This will allow students to learn to use the
toolkit incrementally throughout the course.  Modularity also makes it
easier to change or extend the toolkit.

@@

\noindent\textit{\textbf{Comprehensiveness.}} The toolkit is not intended to
provide a comprehensive set of tools.  Indeed, there should be a wide
variety of ways in which students can extend the toolkit.

\noindent\textit{\textbf{Efficiency.}} The toolkit does not need to be highly
optimized for runtime performance.  However, it should be efficient
enough that students can use their NLP systems to perform real tasks.

\noindent\textit{\textbf{Cleverness.}} Clear designs and implementations are
far preferable to ingenious yet indecipherable ones.

%\section{Python Implementation}
%
%The structure of the Python language directly affects the design and
%implementation of the toolkit.  This section discusses how the design
%of the toolkit is affected by the Python language.
%
%% [EL] This is the only section that I think we might want to keep
%% from python impl.  But I'm not sure we even need to keep this one.
%\noindent\textbf{Interfaces.}
%Python does not directly implement interfaces.  The toolkit
%therefore implements interfaces as simple classes, all of whose methods 
%raise \texttt{AssertionError} exceptions.  A description of the
%functionality provided by the interface is contained in the
%class's documentation string.  Each method's documentation string
%specifies the behavior of that method.
%In order to implement an interface, a class should include that
%interface as one of its bases, and should override every
%(non-optional) method defined by the interface.
%Classes that define interfaces are named with a trailing ``I,'' such
%as \texttt{TokenizerI} or \texttt{EventI}.
%
%% [EL] Mention that Python doesn't provide native type checking?  Or
%% ditch this section? :)
%\noindent\textbf{Typing.}
%Type checking is an important safety device, allowing programmers to
%quickly locate errors in their code.  This is especially important for
%novice programmers, who have less experience with debugging.  Every
%function and method defined by the toolkit will check the types of its
%its arguments.  Given the impact on performance, the level of type checking
%will be adjustable.
%
%\noindent\textbf{Advanced Language Features.}
%The use of advanced language features and programming styles could
%potentially make the toolkit much more difficult to use or to
%understand.  The fewer language features that a student must learn in
%order to use the toolkit, the faster they can start developing NLP
%systems.  The language features used by
%the toolkit are summarized by figure \ref{fig:feature}.
%
%\begin{figure}
%\begin{centering}
%\begin{tabular}{|l|l|}
%\hline
%\noindent\textbf{Language Feature} & \textbf{Use} \\
%\hline
%Classes & used \\
%Exceptions & error conditions only \\
%Operator Overloading & used \\
%Python 2.0 Features & used \\
%Python 2.1 Features & not used \\
%Function Arguments & minimal use \\
%Lambda Functions & minimal use \\
%Mapping and Filtering & not used \\
%List Comprehensions & used \\
%Tkinter & used \\
%\hline
%\end{tabular}\\
%\end{centering}
%\caption{Advanced language features used by NLTK.}
%\label{fig:feature}
%\end{figure}

\section{Modules}
% [EL]: I think that this section works better as a textual
% description, rather than a long list.  Also, we're going to be short
% on space.. :)
% What order should these be in?  Put more impressive stuff at the
% beginning and end?

@@

A set of core modules defines basic data types and processing tasks
that are used througout the toolkit.  The \texttt{token} module
defines basic classes for processing individual elements of text, such
as words or sentences.  The \texttt{tree} module provides data
structures for representing tree strucutres over text, such as syntax
trees and morphological trees.  The \texttt{probability} module
defines classes that encode frequency distributions and probability
distributions, including a variety of smoothing techniques.

@@

The \texttt{parser} module defines a high-level interface for
producing trees that represent the structure of texts.  The
\texttt{chunkparser} module defines a subinterface for parsers that
identify non-overlapping linguistic groups (such as noun phrases) in
unrestricted text.

@@

The \texttt{classifier} module defines a standard interface for
classifying texts into categories.  This interface is currently
implemented by two modules: the \texttt{classifier.naivebayes} module
defines a text classifier based on the Naive Bayes assumption; and the
\texttt{classifier.maxent} module defines the maximum entropy model
for text classification, and includes two algorithms for training the
model: Generalized Iterative Scaling and Improved Iterative Scaling.

@@

%% Old content for this section:
%NLTK currently includes the following modules.
%
%\subsection*{Basics}
%
%\begin{description}
%\item[token:] Basic classes for processing individual
%      elements of text, such as words or sentences. 
%\item[tree:] Classes for representing tree structures
%      over text (such as syntax trees and morphological trees).
%\item[probability:] Classes that encode frequency
%      distributions and probability distributions.
%\end{description}
%
%\subsection*{Tagging}
%
%\begin{description}
%\item[tagger:] A standard interface to tag each token of
%      a text with supplementary information, such as its part of
%      speech; and several implementations of that interface.
%\end{description}
%
%\subsection*{Parsing}
%
%\begin{description}
%\item[parser:] A standard interface to produce trees
%      representing the structure of texts.
%\item[chartparser:] A flexible parser implementation
%      that uses a \emph{chart} to record hypotheses about
%      syntactic constituents.
%\item[srparser\_template:] A partial implementation of a
%      shift-reduce parser; completing the implementation was a
%      student exercise.
%\item[chunkparser:] A standard interface for robust
%      parsers used to identify non-overlapping linguistic groups
%      (such as noun phrases) in unrestricted text.
%\item[rechunkparser:] A regular-expression based
%      implementation of the chunk parser interface.
%\end{description}
%
%\subsection*{Text Classification}
%
%\begin{description}
%\item[classifier:] A standard interface for classifying
%      texts into categories.
%\item[classifier.feature:] A standard way of encoding
%      the information used to make classification decisions.
%\item[classifier.naivebayes:] A text classifier
%      implementation based on the Naive Bayes assumption.
%\item[classifier.maxent:] An implementation of the
%      maximum entropy model for text classification; and
%      implementations of the GIS and IIS algorithms for training
%      the classifier.
%\item[classifier.featureselection:] A standard interface
%      for choosing which features are relevant for making
%      classification decisions.
%\end{description}
%
%\subsection*{Visualization}
%
%\begin{description}
%\item[draw.tree:] A graphical representation for tree
%      structures, such as syntax trees and morphological trees. 
%\item[draw.tree\_edit:] A graphical interface used to
%      build and modify tree structures.
%\item[draw.plot\_graph:] A graphical tool to graph
%      arbitrary functions.
%\item[draw.chart:] An interactive graphical tool used to
%      experiment with chart parsers.
%\end{description}

\section{Uses of NLTK}

\subsection{Assigments and projects}

@@

% Akward
Here we illustrate the use of NLTK in an assignment of moderate
difficulty, for the case of chunk parsing.  This is a kind of parsing
where we only identify the main constituents of a phrase, a technique
with diverse applications.  The NLTK \texttt{chunk} module offers
a variety of rule formats.  For example,
\texttt{ChunkRule('<NN.*>')} builds chunks over consecutive noun tags;
\texttt{ChinkRule('<VB.>')} excises verbs from existing chunks;
\texttt{SplitRule('<NN>', '<DT>')} splits into two pieces any existing
chunk that contains a noun followed by a determiner; and
\texttt{MergeRule('<JJ>', '<JJ>')} combines two adjacent chunks
where the first chunk ends and the second chunk starts with \texttt{JJ}.
The goal of the assignment is to come up with a cascade of these rules,
and to create a set of chunks that are as similar as possible to the test set.

The chunking tutorial provides all the necessary code except for the
chunk rules.  The provided code is responsible for loading the
chunked, tagged text using an existing tokenizer, creating an
unchunked version of the text, applying the chunk rules to the
unchunked text, and scoring the result.  Students focus on the
NLP task only -- providing a ruleset with the best coverage.

@@

\begin{sv}
cascade = [
  ChunkRule('<DT><NN.*><VB.><NN.*>'),
  ChunkRule('<DT><VB.><NN.*>'),
  ChunkRule('<.*>'),
  UnChunkRule('<IN|VB.*|CC|MD|RB.*>'),
  UnChunkRule("<,|{\textbackslash}.|``|''>"),
  MergeRule('<NN.*|DT|JJ.*|CD>', '<NN.*|DT|JJ.*|CD>'),
  SplitRule('<NN.*>', '<DT|JJ>')
]
\end{sv}

This next example illustrates a brute-force approach.  All tags most
commonly found inside of chunks are chunked, while other tags are excluded.

\begin{sv}
cascade = [
  ChunkRule('<{\textbackslash}\$|CD|DT|EX|PDT
             |PRP.*|WP.*|{\textbackslash}\#|FW
             |JJ.*|NN.*|POS|RBS|WDT>*')
]
\end{sv}

The third example involves putting everything in a single chunk,
then excising elements that do not belong.

@@

%% [EL] I'm not sure this is necessary:
%The main program which surrounds the student's work is moderately
%simple as well.  The line replaced by students is underlined.
%% [EL] I find the undeline hard to read; change to italics or bold or
%% vbar in the margin?  Or maybe a \Rightarrow of some sort in the margin?
%\begin{sv}
%from nltk.token import *
%from nltk.chunkparser import *
%from nltk.rechunkparser import *
%import os
%
%path = "/cis530/data/wsj/tagged/"
%
%sent_tok = RETokenizer('(=======+)?{\textbackslash}{\textbackslash}n{\textbackslash}{\textbackslash}n+', 0)
%sentences = []
%files = os.listdir(path)
%for file in files:
%    print 'Loading \%s...' \% file
%    text = open(path+file).read()
%    sentences += sent_tok.tokenize(text, source=file, unit='s')
%
%chunkscore = ChunkScore()
%
%\underline{cascade = [ ??? ]}
%chunkparser = REChunkParser(cascade)
%
%ctt = ChunkedTaggedTokenizer()
%for s in sentences:
%    correct = ctt.tokenize(s.type(), source=s.loc())
%    unchunked = unchunk(correct)
%    guess = chunkparser.parse(unchunked)
%    chunkscore.score(correct, guess)
% 
%print chunkscore
%\end{sv}

The score for each project is the precision, recall and F-measure.
Figure~\ref{fig:contest} shows the results; students with the best two
scores were presented with prizes.

\begin{figure}
\centerline{\epsfig{figure=contest.ps,width=\linewidth}}
\caption{Precision/Recall Graph for Chunking Competition}\label{fig:contest}
\vspace*{2ex}\hrule
\end{figure}

@@

NLTK provides graphical tools that can be used to give in-class
demonstrations, which help to explain basic NLP concepts and
algorithms.  These interactive tools can be used to display relevant
data structures, and to show step-by-step executions of algorithms.
Both data structures and control flow can be easily modified during
the demonstration, in response questions from the class.

Since these graphical tools are included with the toolkit, they can
also be used by students.  This allows students to experiment with the
algorithms that they have seen presented in class.

@@

The chart parsing tool can be used to explain the basic concepts
behind chart parsing, and to show how the algorithm works.  Chart
parsing is a flexible parsing algorithm that uses a data structure
called a \emph{chart} to record hypotheses about syntactic
constituents.  Each hypothesis is represented by a single \emph{edge}
on the chart.  A set of \emph{rules} determine when new edges can be
added to the chart.  This set of rules controls the overall behavior
of the parser (e.g., whether it parses top-down or bottom-up).

@@

This tool can be used to explain several different aspects of chart
parsing.  First, it can be used to explain the basic chart data
structure, and to show how edges can represent hypotheses about
syntactic constituents.  It can then be used to demonstrate and
explain the individual rules that the chart parsers uses to create new
edges.  Finally, it can be used to show how these individual rules
combine to find a complete parse for the sentence.

@@

\begin{figure}
%\centerline{\epsfig{figure=chart.ps,width=\linewidth}}
\caption{Chart Parsing Tool}\label{fig:chartparse}
\vspace*{2ex}\hrule
\end{figure}

\section{Evaluation}

@@

\section{Other approaches}

Java: \cite{Hammond02}
-- no obligation to cite this as it is unpublished
(and I haven't seen a copy).

\section{Conclusion}

@@

>>>>

\documentclass[11pt]{article}
\usepackage{acl02sub,url,alltt,epsfig}
\title{NLTK: The Natural Language Toolkit}
\author{
Edward Loper and Steven Bird\\
Department of Computer and Information Science and Linguistic Data Consortium\\
University of Pennsylvania, Philadelphia, PA 19104-6389, USA
}
\summary{%
NLTK, the Natural Language Toolkit, is a suite of program modules,
tutorials and problem sets, providing ready-to-use computational
linguistics courseware.  NLTK covers symbolic and statistical natural
language processing, and is interfaced to annotated corpora.  Students
augment and replace existing components, learning structured
programming by example, and manipulating sophisticated models from the
outset.}

\paperid{Pxxxx}
\keywords{teaching, courseware, Python, corpora}
\contact{Edward Loper}
\conference{this paper has not been submitted to any other conferences}
\date{\today}

@@

% To do:
%   - Write up evaluation
%   - Write up other approaches
%   - Write up conclusion
%   - General clean-up
%   - ispell
%   - Should URLs include http:// or not (we should be consistant)?

% Outline:
%   - abstract
%   - introduction
%   - choice of programming language
%   - design criteria
%   - modules
%   - uses
%     - assignments/projects
%     - class demonstrations
%   - evaluation
%   - other approaches
%   - conclusion

@@

% ===================== Abstract =====================
% [EL] I think that this abstract should be left out, or shortened and
% merged with the \summary.  The text in the 1st paragraph of the intro
% seems much more direct and relevant to our project than the 1st
% paragraph of this abstract (in particular, the 1st sentence seems like
% a strange way to start). :) I don't think that the second paragraph
% needs to be included in the abstract -- it's not a central point of
% our paper; and the third paragraph is almost identical to the \summary
% (but not quite).  For now, I'm just commenting out the abstract.

%\begin{abstract}
%Students in computational linguistics courses must often learn a new
%programming language.  In some cases, such as when courses are offered
%in linguistics departments, the students may be learning to program
%for the very time.  In order to do interesting projects, it is usually
%necessary for students to do many low-level ``housekeeping'' tasks.
%At the same time, teachers of computational linguistics courses
%sometimes feel that they spend too much time teaching students to
%program, and not enough time teaching the subject itself.  They may
%even avoid programming assignments altogether.  However, we believe
%that it is crucial for a first computational linguistics course to
%include a strong practical component, in which students develop real
%programs to solve real problems with real data.
%
%Python is a new object-oriented scripting language which runs on all
%platforms.  Python has been praised as ``executable pseudocode,'' since
%programs are so easy to write.  Recently, we have been developing
%NLTK, an open-source Natural Language Toolkit written in Python.
%In this presentation, we will motivate, describe and demonstrate NLTK.
%
%NLTK, the Natural Language Toolkit, is a suite of program modules,
%tutorials and problem sets.  NLTK covers symbolic and statistical
%natural language processing, and is interfaced to annotated corpora.
%Students augment and replace existing NLTK components, learning
%structured programming by example, and manipulating sophisticated
%models from the outset.  Along with extensive documentation and
%problem sets, NLTK provides self-contained, ready-to-use CL
%courseware.
%\end{abstract}

% ===================== Introduction =====================
\section{Introduction}

Teachers of introductory courses on computational linguistics are
often faced with the challenge of setting up a practical programming
component for student assignments and projects.  This is a difficult
task because different comptational linguistics domains require a
variety of different data structures and functionality; and because a
diverse range of topics may need to be included in the syllabus.

A widespread practice is to employ multiple programming languages,
where each language provides native data structures and functions that
are a good fit for the task at hand.  For example, a course might use
Prolog for parsing, Perl for corpus processing, and a finite-state
toolkit such for morphological analysis.  By relying on the built-in
features of various languages, the teacher avoids having to develop a
lot of software infrastructure.

An unfortunate consequence is that a significant part of the course
must be devoted to teaching these languages.  Further, many
interesting projects span a variety of domains, and would require that
multiple languages be bridged.  For example, a student project that
involved syntactic parsing of corpus data from a morphologically rich
language might involve all three of the languages mentioned above:
Perl for string processing; a finite state toolkit for morphological
analysis; and Prolog for parsing.
% [EL] This was more precise, but harder to read/parse:
%    (Perl for file I/O, format conversions and
%    overall program control, with calls out to a finite state toolkit for
%    morphological analysis and to a Prolog engine for parsing).  
It is clear that these considerable overheads and shortcomings warrant
a fresh approach.

% [EL] ``difficulties'' doesn't seem like the right word here?
Apart from the practical component, computational linguistics courses
may also depend on software for in-class demonstrations.  This context
calls for highly interactive graphical user interfaces, making it
possible to view program state (e.g. the chart of a chart parser),
observe program execution step-by-step (e.g. execution of a
finite-state machine), and even make minor modifications to programs
in response to ``what if'' questions from the class.  Because of these
difficulties it is common to avoid live demonstrations, and keep
classes for theoretical presentations only.  Apart from being dull,
this approach leaves students to solve important practical problems on
their own, or to deal with them less efficiently in office hours.

% [EL] akward
In this paper we describe a new approach to the above challenges, a
streamlined and flexible way of organizing the practical component of
an introductory computational linguistics course.  We describe NLTK,
the Natural Language Toolkit, which we have developed in conjunction
with a course we have taught at the University of Pennsylvania.

% Overview
% [EL] Given that we're short on space, I decided that this wasn't
% really necessary for a 6-page paper.  If we do include it, it needs
% to be completed (I only went up through sec:uses).
%%Section \ref{sec:python} discusses why we decided to implement the
%%toolkit using Python.  Section \ref{sec:criteria} describes several
%%important considerations and desiderata that guided the design and
%%implementation of the toolkit.  Sections \ref{sec:modules} and
%%\ref{sec:documentation} describe the current contents of the toolkit.
%%Section \ref{sec:uses} explains how the toolkit can be used for both
%%assignments and class demonstrations.

@@

% ===================== Python =====================
\section{Choice of Programming Language}
\label{sec:python}

The most basic step in setting up a practical component is choosing a
suitable programming language.  A number of desiderata influenced our
choice.  First, the language must have a shallow learning curve, so
that novice programmers get immediate rewards for their efforts.
Second, the language must support rapid prototyping and a short
develop/test cycle; an obligatory compilation step is a serious
detraction.  Third, the code should be self-documenting, with a
transparent syntax and semantics.  Fourth, it should be easy to write
structured programs, ideally object-oriented but without the burden
associated with languages like C++.  Finally, the language must have
an easy-to-use graphics library to support the development of simple
graphical user interfaces.

% I moved www.python.org to the \cite.
In surveying the available languages, we believe that Python offers an
especially good fit to the above requirements.  Python is an
object-oriented scripting language developed by Guido van Rossum
and available on all platforms \cite{python}.  Python offers a
shallow learning curve; it was designed to be easily learnt by
children \cite{rossum99}.  As an interpreted language, Python is
suitable for rapid prototyping.  Python code is exceptionally
readable, and it has been praised as ``executable pseudocode.''
% Couldn't find a REF for ``executable pseudocode'' -- closest I got
% was http://www.thinkware.se/cgi-bin/thinki.cgi/PythonQuotes
Python is an object-oriented language, but not punitively so, and it
is easy to encapsulate data and methods inside Python classes.
Finally, Python has an interface to the ``Tk'' graphics toolkit
\cite{tkinter}, and writing graphical interfaces is straightforward.

% ===================== Design Criteria =====================
\section{Design Criteria}
\label{sec:criteria}

Several criteria were considered in the design and implementation of
the toolkit.  The design criteria are divided into primary and
secondary criteria, and listed in the order of their importance.  It
was also important to decide what goals the toolkit would \emph{not}
attempt to accomplish; we therefore include an explicit set of
non-requirements, which the toolkit is not expected to satisfy.

@@

\paragraph{\textit{Ease of Use.}} The primary purpose of the toolkit is
to allow students to concentrate on building NLP systems.  The more
time students must spend learning to use the toolkit, the less useful
it is.

\paragraph{\textit{Consistency.}} The toolkit should use consistent data
structures and interfaces.

\paragraph{\textit{Extensibility.}} The toolkit should easily
accommodate new components, whether those components replicate or
extend the toolkit's existing functionality.  Thus, the toolkit's
design should be modular, with simple and well-defined interfaces
between modules.  The toolkit should also be structured in such a way
that it is obvious where new extensions would fit into the toolkit's
infrastructure.

\paragraph{\textit{Documentation.}} The toolkit, its data structures,
and its implementation all need to be carefully and thoroughly
documented.  Three documentation types are necessary: tutorials, to
teach students how to use the toolkit to perform specific tasks;
technical reports, to explain the toolkit's design and implementation;
and reference documentation, to describe every module, interface,
class and method.  All nomenclature must be carefully chosen and
consistently used.

@@

\paragraph{\textit{Simplicity.}} The toolkit should structure the
complexities of building NLP systems, not hide them.  Therefore, each
class defined by the toolkit should be simple enough that a student
could implement it by the time they finish an introductory course in
computational linguistics.

\paragraph{\textit{Modularity.}} The interaction between different
components of the toolkit should be kept to a minimum.  In particular,
it should be possible to complete individual projects using small
parts of the toolkit, without worrying about how they interact with
the rest of the toolkit.  This allows students to learn how to use the
toolkit incrementally throughout a course.  Modularity also makes it
easier to change and extend the toolkit.

@@

\paragraph{\textit{Comprehensiveness.}} The toolkit is not intended to
provide a comprehensive set of tools.  Indeed, there should be a wide
variety of ways in which students can extend the toolkit.

\paragraph{\textit{Efficiency.}} The toolkit does not need to be highly
optimized for runtime performance.  However, it should be efficient
enough that students can use their NLP systems to perform real tasks.

\paragraph{\textit{Cleverness.}} Clear designs and implementations are
far preferable to ingenious yet indecipherable ones.

% ===================== Modules =====================
\section{Modules}
\label{sec:modules}
% What order should these be in?  Put more impressive stuff at the
% beginning and end?

@@

A set of core modules defines basic data types and processing systems
that are used througout the toolkit.  The \texttt{token} module
defines basic classes for processing individual elements of text, such
as words or sentences.  The \texttt{tree} module provides data
structures for representing tree structures over text, such as syntax
trees and morphological trees.  The \texttt{probability} module
defines classes that encode frequency distributions and probability
distributions, including a variety of statistical smoothing
techniques.

@@

The \texttt{parser} module defines a high-level interface for
producing trees that represent the structures of texts.  The
\texttt{chunkparser} module defines a subinterface for parsers that
identify non-overlapping linguistic groups (such as noun phrases) in
unrestricted text.

@@

The \texttt{classifier} module defines a standard interface for
classifying texts into categories.  This interface is currently
implemented by two modules: the \texttt{classifier.naivebayes} module
defines a text classifier based on the Naive Bayes assumption; and the
\texttt{classifier.maxent} module defines the maximum entropy model
for text classification, and implements two algorithms for training
the model: Generalized Iterative Scaling and Improved Iterative
Scaling.

@@

% ===================== Documentation =====================
\section{Documentation}
\label{sec:documentation}

The toolkit is accompanied by extensive documentation that explains
the toolkit, and describes how to use and extend it.  This
documentation is divided into three primary categories:

\paragraph{\textit{Tutorials}} teach students how to use the toolkit,
in the context of performing specific tasks.  Each tutorial focuses on
a single domain, such as tagging, probablistic systems, or text
classification.  The tutorials include a high-level discussion that
explains and motivates the domain, followed by a detailed
walk-through that uses examples to show how NLTK can be used to
perform specific tasks.

\paragraph{\textit{Reference Documentation}} provides precise
definitions for every module, interface, class, method, function, and
variable in the toolkit.  It is automatically extracted from docstring
comments in the Python source code, using Epydoc \cite{epydoc}.

\paragraph{\textit{Technical Reports}} explain and justify the
toolkit's design and implementation.  They are used by the developers
of the toolkit to guide and document the toolkit's construction.
Students can also consult these reports if they would like further
information about how the toolkit is designed and why it is designed
that way.

% ===================== Uses =====================
\section{Uses of NLTK}
\label{sec:uses}

\subsection{Assigments and Projects}

@@

As an example of a moderately difficult assignment, we asked students
to construct a chunk parser that correctly identifies base noun phrase
chunks in a given text, by defining a cascade of transformational
chunking rules.  The NLTK \texttt{rechunkparser} module provides a
variety of rule types, which the students can instantiate to construct
complete rules.  For example, \texttt{ChunkRule('<NN.*>')} builds
chunks from sequences of consecutive nouns; \texttt{ChinkRule('<VB.>')}
excises verbs from existing chunks; \texttt{SplitRule('<NN>', '<DT>')}
splits any existing chunk that contains a singular noun followed by
determiner into two pieces; and \texttt{MergeRule('<JJ>', '<JJ>')}
combines two adjacent chunks where the first chunk ends and the second
chunk starts with adjectives.

The chunking tutorial motivates chunk parsing, describes each rule
type, and provides all the necessary code for the assignment.  The
provided code is responsible for loading the chunked, part-of-speech
tagged text using an existing tokenizer, creating an unchunked version
of the text, applying the chunk rules to the unchunked text, and
scoring the result.  Students focus on the NLP task only -- providing
a ruleset with the best coverage.

@@

\begin{sv}
cascade = [
  ChunkRule('<DT><NN.*><VB.><NN.*>'),
  ChunkRule('<DT><VB.><NN.*>'),
  ChunkRule('<.*>'),
  UnChunkRule('<IN|VB.*|CC|MD|RB.*>'),
  UnChunkRule("<,|{\textbackslash}{\textbackslash}.|``|''>"),
  MergeRule('<NN.*|DT|JJ.*|CD>', '<NN.*|DT|JJ.*|CD>'),
  SplitRule('<NN.*>', '<DT|JJ>')
]
\end{sv}

The next example illustrates a brute-force statistical approach.  The
student calculated how often each part-of-speech tag was included in a
noun phrase.  They then constructed chunks from any sequence of tags
that occured in a noun phrase more than 50\% of the time.

\begin{sv}
cascade = [
  ChunkRule('<{\textbackslash}{\textbackslash}\$|CD|DT|EX|PDT
             |PRP.*|WP.*|{\textbackslash}{\textbackslash}\#|FW
             |JJ.*|NN.*|POS|RBS|WDT>*')
]
\end{sv}

In the third example, the student constructed a single chunk
containing the entire text, and then excised all elements that did not
belong.

@@

% [EL] This makes it sound like we GRADED them using their precision,
% recall, and f-measure.  Either say something about the contest, or
% leave this out:
Each student's project was scored using its precision, recall, and
F-measure.

% [EL] Leave out the picture.
%Figure~\ref{fig:contest} shows the results; students with the best two
%scores were presented with prizes.
%
%\begin{figure}
%\centerline{\epsfig{figure=contest.ps,width=\linewidth}}
%\caption{Precision/Recall Graph for Chunking Competition}\label{fig:contest}
%\vspace*{2ex}\hrule
%\end{figure}

@@

NLTK provides graphical tools that can be used to give in-class
demonstrations that help to explain basic NLP concepts and algorithms.
These interactive tools can be used to display relevant data
structures, and to show the step-by-step execution of algorithms.
Both data structures and control flow can be easily modified during
the demonstration, in response questions from the class.

Since these graphical tools are included with the toolkit, they can
also be used by students.  This allows students to experiment at home
with the algorithms that they have seen presented in class.

@@

The chart parsing tool is an example of a graphical tool provided by
NLTK.  This tool can be used to explain the basic concepts behind
chart parsing, and to show how the algorithm works.  Chart parsing is
a flexible parsing algorithm that uses a data structure called a
\emph{chart} to record hypotheses about syntactic constituents.  Each
hypothesis is represented by a single \emph{edge} on the chart.  A set
of \emph{rules} determine when new edges can be added to the chart.
This set of rules controls the overall behavior of the parser (e.g.,
whether it parses top-down or bottom-up).

@@

This tool can be used to explain several different aspects of chart
parsing.  First, it can be used to explain the basic chart data
structure, and to show how edges can represent hypotheses about
syntactic constituents.  It can then be used to demonstrate and
explain the individual rules that the chart parsers uses to create new
edges.  Finally, it can be used to show how these individual rules
combine to find a complete parse for a given sentence.

@@

\begin{figure}
\centerline{\epsfig{figure=chartparse.eps,width=\linewidth}}
\caption{Chart Parsing Tool}\label{fig:chartparse}
\vspace*{2ex}\hrule
\end{figure}

% ===================== Evaluation =====================
\section{Evaluation}
\label{sec:evaluation}

@@

% ===================== Other Approaches =====================
\section{Other Approaches}
\label{sec:approaches}

\textit{This section will include a discussion of other approaches to
implementing computational linguistics courseware, including:
Programming for Linguistics: Java Technology for Language
Researchers\cite{Hammond02}; The OpenNLP project\cite{opennlp};
others?}

% ===================== Conclusion =====================
\section{Conclusion}
\label{sec:conclusion}

@@

% include URL (a second time; 1st was in the intro) in conclusion
